A Spectral Analysis of the Domain Decomposed Adjoint Neumann-Ulam Method

### 2012-11-27 doc/mandc2013/notes/spectral_theory.tex
fixed optical thickness reltationships
adding additional spectral data

In the case of domain decomposed linear operator equations, the chord
length can be interpreted as the number of discrete states spanning a
domain, $n_i$, along the typical length of that domain (e.g. the
number of discrete states in the $i$ direction in a square domain
using our model diffusion problem). We then seek the equivalent of the
total cross section, $\Sigma_t$ in terms of the discrete random walk
behavior. The total cross section can also be interpreted as the
inverse of the mean free path of particles in the system. In the
Neumann-Ulam method, this is analogous to the length of the Markov
chains (and therefore the number of state-to-state transitions)
generated while sampling the Neumann series. In the previous section,
we determined this length with Eq~(\ref{eq:analytic_k}) to be
dependent on the Eigenvalues of the operator. In addition, we must
also consider the number of states that may be traversed in a given
transition, $n_s$, along the path of the typical length of the
system. For our diffusion model problem, $n_s$ would equate to the
number of states in the $i$ (or $j$ as the problem is symmetric)
direction that a history could potentially move and is dependent on
the stencil used for the discretization. We then have the following
relation for the \textit{optical thickness}, $\tau$, of a domain in
the decomposition:
\begin{equation}
  \tau = \frac{n_i n_s}{k}\:,
  \label{eq:matrix_optical_thickness}
\end{equation}
and inserting Eq~(\ref{eq:analytic_k}) gives:
\begin{equation}
  \tau = \frac{n_i n_s \log(\rho(\ve{H}))}{\log(W_c)}\:.
  \label{eq:matrix_optical_thickness_2}
\end{equation}
Using this optical thickness, we can then complete the Wigner rational
approximation by defining the bounds of $\tau \rightarrow 0, \lambda
\rightarrow 1$ and $\tau \rightarrow \infty, \lambda \rightarrow
\tau^{-1}$. We expect this behavior as optically thin domains will
communicate most of their histories while optically thick domains will
leak the fraction of histories that did not interact within. Using
the Wigner approximation, we can then define the leakage fraction out
of a domain for the adjoint Neumann-Ulam method:
\begin{equation}
  \lambda = \frac{1}{1+\tau}\:.
  \label{eq:analytic_domain_leakage}
\end{equation}
Here, the leakage fraction is explicitly bound to the Eigenvalues of
the iteration matrix, the size of the domain, the content of the
discretization stencil, and the weight cutoff selected to terminate
low weight histories.  We also note here an interesting consequence of
this approximation: no matter how optically thick a domain may be, the
domain is still expected to leak some histories and thus communicate
them. Therefore, no matter how large the domains are made, if they do
not encapsulate the entire global domain they require some
communication in order to implement a domain decomposed adjoint
Neumann-Ulam method.

>>>>

In the case of domain decomposed linear operator equations, the chord
length can be interpreted as the number of discrete states spanning a
domain, $n_i$, along the typical length of that domain (e.g. the
number of discrete states in the $i$ direction in a square domain
using our model diffusion problem). We then seek the equivalent of the
total cross section, $\Sigma_t$ in terms of the discrete random walk
behavior. The total cross section can also be interpreted as the
inverse of the mean free path of particles in the system. In the
Neumann-Ulam method, this is analogous to the length of the Markov
chains (and therefore the number of state-to-state transitions)
generated while sampling the Neumann series. In the previous section,
we determined this length with Eq~(\ref{eq:analytic_k}) to be
dependent on the Eigenvalues of the operator. In addition, we must
also consider the number of states that may be traversed in a given
transition, $n_s$, along the path of the typical length of the
system. For our diffusion model problem, $n_s$ would equate to the
number of states in the $i$ (or $j$ as the problem is symmetric)
direction that a history could potentially move in a single transition
and is dependent on the stencil used for the discretization. We then
have the following relation for the \textit{optical thickness},
$\tau$, of a domain in the decomposition:
\begin{equation}
  \tau = \frac{n_i}{n_s k}\:,
  \label{eq:matrix_optical_thickness}
\end{equation}
and inserting Eq~(\ref{eq:analytic_k}) gives:
\begin{equation}
  \tau = \frac{n_i \log(\rho(\ve{H}))}{n_s \log(W_c)}\:.
  \label{eq:matrix_optical_thickness_2}
\end{equation}
Using this optical thickness, we can then complete the Wigner rational
approximation by defining the bounds of $\tau \rightarrow 0, \lambda
\rightarrow 1$ and $\tau \rightarrow \infty, \lambda \rightarrow
\tau^{-1}$. We expect this behavior as optically thin domains will
communicate most of their histories while optically thick domains will
leak the fraction of histories that did not interact within. Using
the Wigner approximation, we can then define the leakage fraction out
of a domain for the adjoint Neumann-Ulam method:
\begin{equation}
  \lambda = \frac{1}{1+\tau}\:.
  \label{eq:analytic_domain_leakage}
\end{equation}
Here, the leakage fraction is explicitly bound to the Eigenvalues of
the iteration matrix, the size of the domain, the content of the
discretization stencil, and the weight cutoff selected to terminate
low weight histories.  We also note here an interesting consequence of
this approximation: no matter how optically thick a domain may be, the
domain is still expected to leak some histories and thus communicate
them. Therefore, no matter how large the domains are made, if they do
not encapsulate the entire global domain they require some
communication in order to implement a domain decomposed adjoint
Neumann-Ulam method.
